Inference of proto-neutron star properties from gravitational wave data in core-collapse supernovae.

### 2020-10-13 results.tex
Add EOS comment
add NLC comments

To estimate how accurately we can infer the time evolution of $r=M_{\rm PNS}/R_{\rm PNS}^2$ in the
gravitational wave detector data, we have added \pcd{the} GW signal \pcd{from {\texttt s20S}} to 
100 Gaussian noise realisations whose power spectral density follows \pcd{the} advanced LIGO (aLIGO)
spectrum~\cite{aLIGOsens:2018} shown on Figure~\ref{fig:spectrum}.

\begin{figure}
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors advanced LIGO (aLIGO) and Advanced Virgo (ADV) at design sensitivity and of the proposed third generation detectors Cosmic Explorer and Einstein Telescope. Eintein Telescope sensitivity curve ET-B is obtained pushing second generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the “A+” upgrade to advanced LIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. } \label{fig:spectrum}
\end{figure}

We have varied the distance to the source, covering a large
range of distances for which a detection in second generation of gravitational wave detectors
is feasible. The source is optimally oriented with
respect to the gravitational wave detector. We are assuming a GW signal from a core collapse
phenomena has been identified in the data and that the beginning of the GW signal is known within $O(10~ms)$.
The data (signal embedded in noise) are whitened using the function {\it prewhiten} of the R-package TSA.
An auto-regressive model with maximal \textcolor{red}{100} coefficients has been used.

For each of the noise realisations, we reconstruct the ratio time series {$r_i$}
of length $N$ starting from the left side of the spectrogram and constraining the beginning of the
track to be smaller than \unit[200]{Hz}. The reconstructed ratio is then compare to the ``true'' ratio
{$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation.

Figure \ref{fig:s20results} is showing the distribution of the fraction of the ratio {$r_i^0$} values
that fall within the 95\% confidence interval of {$r_i$}. This quantity, $coverage$, is taking maximal values
when the source is located within few kpc and then decreases with the distance.

@@

$\Delta$ values of each of the 100 noise realisation are shown as well as function of the distance
on Figure \ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%.
At small distance, $\Delta$ is small but not null. This reflects the approximation of the model used for $r$.
It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision at distance up to $\sim$ 9 kpc for this particular waveform, with $coverage$ value
larger than 80\%. There are few noise realisations for a source located at $<$ \unit[9]{kpc} for which
$\Delta$ takes large values, indicating that the method start failing to reconstruct with accuracy the ratio.

@@

 \caption{Boxplots of $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realisations is considered for each distance.}
  \label{fig:s20results}
\end{figure}

@@

Figure \ref{fig:aLIGO_cov_allwvf} shows that apart {\tt s11.2--LS220} and to a lesser extent
{\tt s20.0--SFHo}, the ratio is well reconstructed for all waveforms up to $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio
we have run 100 simulations without injecting a signal and have measured $coverage$ for the
reconstructed ratios.
The median of $coverage$ as well as the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is null in this case, but it can be different from zero because
the g-mode reconstruction algorithm is looking for a continuously frequency increasing track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effects explains why outliers can reach
values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered. 
In Table \ref{tab:results} we are reporting the distance $d_r$ at which $coverage$ median is lower
than 95\% of the noise only values. We have checked that $coverage$ and $\Delta$ provide similar
values.
These numbers are an estimate of the order of magnitude of the source maximal distance at which a
reconstruction of the ratio could be possible with current GW detectors.
They are also upper limits as we are taking into account the detector antenna response in our
simulation but consider the source is optimally oriented. 
Table \ref{tab:results} reports also $d_r$ for the advanced Virgo detector at design sensitivity.
Results are very similar to aLIGO, despite the detector sensitivity differences.
Note that Table \ref{tab:results} provides the distance at which one could detect a source
optimally oriented with a matched filter signal-to-noise ratio of 13.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
 \caption{Median of $Delta$ for 8 CCSN waveforms embedded in aLIGO noise and located at different distance from the Earth. The ``no signal'' line and band show the median and first and third quartile of $Delta$ in absence of any signal.} \label{fig:aLIGO_prec_allwvf}
\end{figure}

The same analysis has been performed using expected sensitivity curves for the third generation of
gravitational wave detectors. In Europe the Einstein Telescope project proposes to host in a 10-km
equilateral triangle configuration 3 low-power low-frequency cryogenic interferometers as well as 3
high-power high-frequency interferometers. Three sensitivity curves, ET-B, ET-C and ET-D corresponding
to different options and stages of the project~\cite{Hild_2011} are considered in this study.
The US based project Cosmic Explorer~\cite{reitze2019cosmic} is proposing to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2 also shown in Figure~\ref{fig:spectrum}.

Figure \ref{fig:s20--SFHo_all3G} shows $\Delta$ as function of the source distance for {\tt s20.0--SFHo}
waveform for the 5 3G detectors configurations. Overall, the ratio is well reconstructed up to distances
in the range \unit{100--200}[kpc] which represents an improved of at least a factor 10 with respect to
advanced LIGO and advanced Virgo detectors. We can also note that the Einstein Telescope results lay in
between the 2 Cosmic Explorer results. This is confirmed for all other waveforms, expect {\tt s25.0-LS220}
for which the maximal distance reach in CE2 is significantly lower than CE1. This is partly due to the
small variation of the reconstruction quality to the distance of the source making the estimation of $d_r$
rather incertain for this waveform. All results are summarized in Table \ref{tab:results} and Figure
\ref{fig:distances}. It is remarkable that with 3G detectors the ratio could be reconstructed for sources
located up to several hundred of kpc. It is nevertheless important to note the rather wide range obtained
for the different waveforms probing a large range of progenitor masses.
We did not find any correlation between the mass of the progenitor and $d_r$. On the other hand, the quality
of the ratio reconstruction depends on the signal-to-noise ratio, expressed in Table \ref{tab:results} by
$d_{det}$.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of $\Delta$ for s20.0--SFHo CCSN waveform embedded in 3G detectors noise and located at different distance from the Earth. } \label{fig:s20--SFHo_all3G}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance $d_{r}$ at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for a source optimally oriented with respect to the GW detectors for the 7 CCSN waveforms considered in this study.} \label{fig:distances}
\end{figure}

@@

  \end{tabular}
  \caption{%%
    Maximal distance $d_{r}$ at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for a source optimally oreinted with respect to the GW detectors
    considered in this study. $d_{det}$ is the distance at which one could detect a source
    optimally oriented with a matched filter signal-to-noise ratio of 13 in the  different
    GW detectors.
    %%
  }
  \label{tab:results}
\end{table}

>>>>

To estimate how accurately we can infer the time evolution of $r=M_{\rm PNS}/R_{\rm PNS}^2$ in a single
detector GW data, we have added the GW signal {\texttt s20S} to 
100 Gaussian noise realisations whose power spectral density follows the Advanced LIGO (aLIGO)
spectrum~\cite{aLIGOsens:2018} shown on Figure~\ref{fig:spectrum}.

\begin{figure}
 \centering
 \includegraphics[width=0.5\textwidth]{plots/spectrum}
 \caption{Amplitude spectral density of the GW detectors Advanced LIGO (aLIGO) and Advanced Virgo (ADV) at design sensitivity and of the proposed third generation detectors Cosmic Explorer and Einstein Telescope. Eintein Telescope sensitivity curve ET-B is obtained pushing second generation detector technology at its limit. ET-C and ET-D sensitivity curves correspond to a detector configuration where a low-power cryogenic low-frequency interferometer and a high-power room temperature high-frequency interferometer are sharing the same infrastructure~\cite{Hild_2011}. Cosmic Explorer design sensitivity will be achieved in two stages. Stage 1 (CE1) is expected to use the technology developed for the “A+” upgrade to Advanced LIGO but scaled up to a 40 km detector while stage 2 (CE2) will implement state-of-the-art technology to decrease quantum and thermal noises~\cite{reitze2019cosmic}. } \label{fig:spectrum}
\end{figure}

We have covered a large range of distances for which a detection in second generation of GW
detectors is feasible. The source is optimally oriented with
respect to the GW single detector. We are assuming a GW signal from a core collapse
phenomena has been identified in the data and that the beginning of the GW signal is known within $O(10~ms)$.
The data (signal embedded in noise) are whitened using the function {\it prewhiten} of the R-package TSA.
An auto-regressive model with maximal \textcolor{red}{100} coefficients has been used.

For each of the noise realizations, we reconstruct the ratio time series {$r_i$}
of length $N$ starting from the left side of the spectrogram and constraining the beginning of the
track to be smaller than \unit[200]{Hz}. The reconstructed ratio is then compared to the ``true'' ratio
{$r_i^0$} derived from the PNS mass and radius computed from the {\texttt s20S} simulation.

Figure \ref{fig:s20results} shows the distribution of the fraction of the ratio {$r_i^0$} values
that fall within the 95\% confidence interval of {$r_i$}. This quantity, $coverage$, is taking maximal values
when the source is located within few kpc and then decreases with the distance.

@@

$\Delta$ values of each of the 100 noise realisation are shown as well as function of the distance
on Figure \ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%.
At small distances, $\Delta$ is small but not null. This reflects the approximation of the model used for $r$.
It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision at distance up to $\sim$ 9 kpc for this particular waveform, with $coverage$ value
larger than 80\%. There are few noise realizations for a source located at $<$ \unit[9]{kpc} for which
$\Delta$ takes large values, indicating that the method start failing to reconstruct with accuracy the ratio.

@@

 \caption{Boxplots of the $coverage$ (upper panel) and $\Delta$ (lower panel) for {\texttt s20S} signal embedded in aLIGO noise at different distances from the Earth. 100 noise realizations are considered for each distance.}
  \label{fig:s20results}
\end{figure}

@@

Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from {\texttt s11} and to a lesser extent
{\texttt s20S}, the ratio is well reconstructed for all waveforms up to $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio
we have run 100 simulations without injecting a signal and have measured $coverage$ for the
reconstructed ratios.
The median of $coverage$ as well as the 95 quartile are shown on Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is null in this case, but it can be different from zero because
the g-mode reconstruction algorithm is looking for a continuously frequency increasing track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effects explains why outliers can reach
values as high as 80\%.
Figure \ref{fig:aLIGO_prec_allwvf} shows $\Delta$ as function of the distance for the same signals
as well as the result when only noise is considered. 
In Table \ref{tab:results} we are reporting the distance $d_r$ at which $coverage$ median is lower
than 95\% of the noise only values. We have checked that $coverage$ and $\Delta$ provide similar
values.
These numbers are an estimate of the order of magnitude of the source maximal distance at which a
reconstruction of the ratio could be possible with current GW detectors.
They are also upper limits as we are taking into account the detector antenna response in our
simulation but consider the source is optimally oriented. 
Table \ref{tab:results} reports also $d_r$ for the Advanced Virgo detector at design sensitivity.
Results are very similar to aLIGO, despite the detector sensitivity differences.
Note that Table \ref{tab:results} provides the distance at which one could detect a source
optimally oriented with a matched filter signal-to-noise ratio of 13.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
 \caption{Median of $\Delta$ for 8 CCSN waveforms embedded in aLIGO noise and located at different distance from the Earth. The ``no signal'' line and band show the median and first and third quartile of $\Delta$ in absence of any signal.} \label{fig:aLIGO_prec_allwvf}
\end{figure}

The same analysis has been performed using expected sensitivity curves for the third generation of
GW detectors. In Europe the Einstein Telescope project proposes to host in a 10-km
equilateral triangle configuration 3 low-power low-frequency cryogenic interferometers as well as 3
high-power high-frequency interferometers. Three sensitivity curves, ET-B, ET-C and ET-D corresponding
to different options and stages of the project~\cite{Hild_2011} are considered in this study.
The US based project Cosmic Explorer~\cite{reitze2019cosmic} is proposing to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2 also shown in Figure~\ref{fig:spectrum}.

Figure \ref{fig:s20--SFHo_all3G} shows $\Delta$ as function of the source distance for {\texttt s20S}
waveform for the five 3G detector configurations. Overall, the ratio is well reconstructed up to distances
in the range \unit[100--200]{kpc} which represents an improvement of a factor 10 with respect to
Advanced LIGO and Advanced Virgo detectors. We can also note that the Einstein Telescope results lay in
between the 2 Cosmic Explorer results. This is confirmed for all other waveforms, expect {\texttt s25}
for which the maximal distance reach in CE2 is significantly lower than CE1. This is partly due to the
small variation of the reconstruction quality to the distance of the source making the estimation of $d_r$
rather uncertain for this waveform. All results are summarized in Table \ref{tab:results} and Figure
\ref{fig:distances}. It is remarkable that with 3G detectors the ratio could be reconstructed for sources
located up to several hundred of kpc. It is nevertheless important to note the rather wide range obtained
for the different waveforms probing a large range of progenitor masses.
We did not find any correlation between the mass of the progenitor and $d_r$, nor the equation of state.
On the other hand, the quality
of the ratio reconstruction depends on the signal-to-noise ratio, expressed in Table \ref{tab:results} by
$d_{det}$.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of $\Delta$ for {\texttt s20S} CCSN waveform embedded in 3G detectors noise and located at different distance from the Earth. } \label{fig:s20--SFHo_all3G}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{plots/dist_allwvfs_2G3G}
  \caption{Maximal distance $d_{r}$ in kpc at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for a source optimally oriented with respect to the GW detectors for the 7 CCSN waveforms considered in this study.} \label{fig:distances}
\end{figure}

@@

  \end{tabular}
  \caption{%%
    Maximal distance $d_{r}$ at which the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ is reconstructed
    with good accuracy for a source optimally oreinted with respect to the GW detectors
    considered in this study. $d_{det}$ is the distance at which one could detect a source
    optimally oriented with a matched filter signal-to-noise ratio of 13 in the  different
    GW detectors. All distances are expressed in kpc.
    %%
  }
  \label{tab:results}
\end{table}

### 2020-10-13 simulations.tex
add NLC comments

{Unless other methods used GW astronomy, the} algorithm proposed in {this work} does not require accurate
waveforms {in order to infer the properties of the PNS.} {Instead, it relies on the evolution of the frequency of
oscillation of some particular modes, as seen in the GW spectrum. The frequency of these modes depends, in a universal way, 
on the surface gravity of the PNS ($r=M_{\rm PNS}/R_{\rm PNS}^2$), in the sense that if at a given time we observe GW emission
at a certain frequency $f$ we can determine univocally the value of the surface gravity, within a certain error, regardless of the details
of the numerical simulation. In this work we use two sets of simulations: i) The {\it model set}, composed by 1D simulations, which is used 
to build the universal relation (model), $r(f)$, linking the ratio $r$ with the observed frequency $f$, and ii) the {\it test set}, composed by 2D simulations, 
for which we know both the GW signal and the evolution of the ratio, $r (t)$, and that is used to test performance of the algorithm.}

@@

{For the {\it model set}, we use the $25$ spherically symmetric (1D) simulations of \citep{Torres:2019a}
including progenitors with zero-age main sequence (ZAMS) masses in the range 
$M_{\mathrm{ZAMS}} = 11.2 - 75 \, \Msol$. The set contains simulations using the
 two numerical codes and six different equations of state. Details can be found in
 \citep{Torres:2019a}. The reason to use one dimensional simulations for the model set
 is that the computational cost of those is significantly smaller than the cost of multidimensional
 simulations, so is easier to accumulate the statistics necessary to build a good model for $r(f)$.}
 \pcd{For each time of each simulation we compute the ratio $r$ and the frequency of the $^2g_2$ mode by means of the linear analysis
 described in \cite{Torres:2018,Torres:2019a,Torres:2019b}. }

@@

{For all the simulations,} we mapped the pre-collapse state of the stars to a spherical
coordinate system with $n_r = 400$ zones in radial direction
distributed logarithmically with a minimum grid width of
$(\Delta r)_{\mathrm{min}} = 400 \, \mathrm{m}$ and an outer radius of
$r_{\mathrm{max}} = 8.3 \times 10^{9} \, \mathrm{cm}$ and
$n_{\theta} = 128$ equidistant cells in angular direction. For the
neutrino energies, we used a logarithmic grid with $n_e = 10$ bins up
to $240 \, \mathrm{MeV}$.
\pcd{Unlike the model set, the simulations in the test set are not 1D because we need to 
extract the gravitational wave signal, which is a multi-dimensional effect. For each simulation
we extract the gravitational wave signal, $h_+(t)$,  by means of the quadrupole formula and compute the 
time evolution of the surface gravity, $r(t)$.}

All spherical and most axisymmetric models fail to achieve shock
revival during the time of our simulations. Only the two stars with
the highest masses, \texttt{s25} and \texttt{s40}, develop relatively
late explosions in axisymmetry. Consequently, mass accretion onto the
PNSs proceeds at high rates for a long time in all cases and causes
them to oscillate with their characteristic frequencies. The final
masses of the PNSs are in the range of
$M_{\mathrm{PNS}} = 1.47 - 2.33 \, \Msol$, i.e., insufficient for
producing a black hole.

>>>>

{Unlike other methods used GW astronomy, the} algorithm proposed in {this work} does not require accurate
waveforms {in order to infer the properties of the PNS.} {Instead, it relies on the evolution of the
frequency of oscillation of some particular modes, as seen in the GW spectrum.
The frequency of these modes depends, in a universal way, on the surface gravity of the PNS
($r=M_{\rm PNS}/R_{\rm PNS}^2$), in the sense that if at a given time we observe GW emission at a certain
frequency $f$ we can determine unequivocally the value of the surface gravity, within a certain error,
regardless of the details of the numerical simulation. In this work we use two sets of simulations: i)
The {\it model set}, composed by 1D simulations, which is used to build the universal relation (model),
$r(f)$, linking the ratio $r$ with the observed frequency $f$, and ii) the {\it test set}, composed by
2D simulations, 
for which we know both the GW signal and the evolution of the ratio, $r (t)$, and that is used to test
performance of the algorithm.}

@@

{For the {\it model set}, we use the $25$ spherically symmetric (1D) simulations of \citep{Torres:2019a}
including progenitors with zero-age main sequence (ZAMS) masses in the range 
$M_{\mathrm{ZAMS}} = 11.2 - 75 \, \Msol$. The set contains simulations using the
 two numerical codes and six different equations of state. Details can be found in
 \citep{Torres:2019a}. The reason to use one dimensional simulations for the model set
 is that the computational cost of those is significantly smaller than the cost of multidimensional
 simulations, so is easier to accumulate the statistics necessary to build a good model for $r(f)$.}
 {For each time of each simulation we compute the ratio $r$ and the frequency of the $^2g_2$ mode by means of the linear analysis
 described in \cite{Torres:2018,Torres:2019a,Torres:2019b}. }

@@

{For all the simulations,} we mapped the pre-collapse state of the stars to a spherical
coordinate system with $n_r = 400$ zones in radial direction
distributed logarithmically with a minimum grid width of
$(\Delta r)_{\mathrm{min}} = 400 \, \mathrm{m}$ and an outer radius of
$r_{\mathrm{max}} = 8.3 \times 10^{9} \, \mathrm{cm}$ and
$n_{\theta} = 128$ equidistant cells in angular direction. For the
neutrino energies, we used a logarithmic grid with $n_e = 10$ bins up
to $240 \, \mathrm{MeV}$.
{Unlike the model set, the simulations in the test set are not 1D because we need to 
extract the GW signal, which is a multi-dimensional effect. For each simulation
the GW signal, $h_+(t)$, is extracted by means of the quadrupole formula and we compute the 
time evolution of the surface gravity, $r(t)$.}

All spherical and most axisymmetric models fail to achieve shock
revival during the time of our simulations. Only the two stars with
the highest masses, \texttt{s25} and \texttt{s40}, develop relatively
late explosions in axisymmetry. Consequently, mass accretion onto the
PNSs proceeds at high rates for a long time in all cases and causes
them to oscillate with their characteristic frequencies. The final
masses of the PNSs are in the range of
$M_{\mathrm{PNS}} = 1.47 - 2.33 \, \Msol$, i.e., likely insufficient for
producing a black hole.
